---
title: "Hands-on Exercise 1: Geospatial data Wrangling with R"
editor: visual
---

## Overview

In this hands-on exercise, I learn how to import and wrangling geospatial data using appropriate R packages.

## Getting Started

The code chunk below install and load [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/packages/) packages into R environment.

```{r}
pacman::p_load(sf,tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data in shapefile form

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

### Importing GIS data in kml format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

## Checking the Content of A Simple Feature Data Frame

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

### Working with glimpse()

```{r}
glimpse(mpsz)
```

### 

*glimpse()* report reveals the data type of each fields. For example `FMEL-UPD_D` field is in **date** data type and `X_ADDR`, `Y_ADDR`, `SHAPE_L` and `SHAPE_AREA` fields are all in **double-precision values**.

Working with head()

Sometimes we would like to reveal complete information of a feature object, this is the job of [*head()*](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/head) of Base R

```{r}
head(mpsz, n=5)
```

Note: One of the useful argument of *head()* is it allows user to select the numbers of record to display (i.e. the n argument).

## Plotting the Geospatial Data

In geospatial data science, by looking at the feature information is not enough. We are also interested to visualise the geospatial features. This is the time you will find *plot()* of R Graphic comes in very handy as shown in the code chunk below.

```{r}
plot(mpsz)
```

The default plot of an sf object is a multi-plot of all attributes, up to a reasonable maximum as shown above. We can, however, choose to plot only the geometry by using the code chunk below.

```{r}
plot(st_geometry(mpsz))
```

Alternatively, we can also choose the plot the sf object by using a specific attribute as shown in the code chunk below.

```{r}
plot(mpsz["PLN_AREA_N"])
```

Note: *plot()* is mean for plotting the geospatial object for quick look. For high cartographic quality plot, other R package such as tmap should be used.

## Working with Projection

Map projection is an important property of a geospatial data. In order to perform geoprocessing using two geospatial data, we need to ensure that both geospatial data are projected using similar coordinate system.

In this section, you will learn how to project a simple feature data frame from one coordinate system to another coordinate system. The technical term of this process is called **projection transformation**.

### 1.7.1 Assigning EPSG code to a simple feature data frame

One of the common issue that can happen during importing geospatial data into R is that the coordinate system of the source data was either missing (such as due to missing .proj for ESRI shapefile) or wrongly assigned during the importing process.

This is an example the coordinate system of `mpsz` simple feature data frame by using *st_crs()* of *sf* package as shown in the code chunk below.

```{r}
st_crs(mpsz)
```

Although `mpsz` data frame is projected in svy21 but when we read until the end of the print, it indicates that the EPSG is 9001. This is a wrong EPSG code because the correct EPSG code for svy21 should be [3414](https://epsg.io/3414).

In order to assign the correct EPSG code to `mpsz` data frame, *st_set_crs()* of **sf** package is used as shown in the code chunk below.

```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

Notice that the EPSG code is 3414 now.

### Transforming the projection of preschool from wgs84 to svy21.

In geospatial analytics, it is very common for us to transform the original data from geographic coordinate system to projected coordinate system. This is because geographic coordinate system is not appropriate if the analysis need to use distance or/and area measurements.

Let us take preschool simple feature data frame as an example. The print below reveals that it is in wgs84 coordinate system.

    Geometry set for 1359 features 
    Geometry type: POINT
    Dimension:     XYZ
    Bounding box:  xmin: 103.6824 ymin: 1.248403 xmax: 103.9897 ymax: 1.462134
    z_range:       zmin: 0 zmax: 0
    Geodetic CRS:  WGS 84
    First 5 geometries:

This is a scenario that *st_set_crs()* is not appropriate and *st_transform()* of sf package should be used. This is because we need to reproject `preschool` from one coordinate system to another coordinate system mathemetically.

Let us perform the projection transformation by using the code chunk below.

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

```{r}
st_geometry(preschool3414) 
```

Notice that it is in svy21 projected coordinate system now. Furthermore, if you refer to *Bounding box:*, the values are greater than 0-360 range of decimal degree commonly used by most of the geographic coordinate systems.

## 1.8 Importing and Converting An Aspatial Data

In practice, it is not unusual that we will come across data such as `listing` of Inside Airbnb. We call this kind of data aspatial data. This is because it is not a geospatial data but among the data fields, there are two fields that capture the x- and y-coordinates of the data points.

In this section, you will learn how to import an aspatial data into R environment and save it as a tibble data frame. Next, you will convert it into a simple feature data frame.

For the purpose of this exercise, the `listings.csv` data downloaded from AirBnb will be used.

### Importing the aspatial data

Since `listings` data set is in csv file format, we will use [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package to import `listing.csv` as shown the code chunk below. The output R object is called `listings` and it is a [tibble data frame](https://r4ds.had.co.nz/tibbles.html).

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

After importing the data file into R, it is important for us to examine if the data file has been imported correctly.

The code chunk below shows *list()* of Base R instead of *glimpse()* is used to do the job.

```{r}
list(listings)
```

The output reveals that `listing` tibble data frame consists of 4252 rows and 16 columns. Two useful fields we are going to use in the next phase are `latitude` and `longitude`. Note that they are in decimal degree format. As a best guess, we will assume that the data is in **wgs84** Geographic Coordinate System.

### Creating a simple feature data frame from an aspatial data frame

The code chunk below converts `listing` data frame into a simple feature data frame by using [*st_as_sf()*](https://r-spatial.github.io/sf/reference/st_as_sf.html) of **sf** packages

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

Things to learn from the arguments above:

-   *coords* argument requires you to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.

-   *crs* argument requires you to provide the coordinates system in epsg format. [EPSG: 4326](https://epsg.io/4326) is wgs84 Geographic Coordinate System and [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System. You can search for other country's epsg code by referring to [epsg.io](https://epsg.io/).

-   *%\>%* is used to nest *st_transform()* to transform the newly created simple feature data frame into svy21 projected coordinates system.

Let us examine the content of this newly created simple feature data frame.

```{r}
glimpse(listings_sf)
```

## 1.9 Geoprocessing with sf package

Besides providing functions to handling (i.e. importing, exporting, assigning projection, transforming projection etc) geospatial data, **sf** package also offers a wide range of geoprocessing (also known as GIS analysis) functions.

In this section, you will learn how to perform two commonly used geoprocessing functions, namely [buffering](https://www.gislounge.com/buffers-in-gis/) and point in polygon count.

### 1.9.1 Buffering

The scenario:

The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 metres of reserved land on the both sides of the current cycling path. You are tasked to determine the extend of the land need to be acquired and their total area.

The solution:

Firstly, [*st_buffer()*](https://r-spatial.github.io/sf/reference/geos_unary.html) of **sf** package is used to compute the 5-meter buffers around cycling paths

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

This is followed by calculating the area of the buffers as shown in the code chunk below.

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

Lastly, sum() of Base R will be used to derive the total land involved

```{r}
sum(buffer_cycling$AREA)
```

**Mission Accomplished!**

### 1.9.2 Point-in-polygon count

The scenario:

A pre-school service group want to find out the numbers of pre-schools in each Planning Subzone.

The solution:

The code chunk below performs two operations at one go. Firstly, identify pre-schools located inside each Planning Subzone by using [st_intersects()](https://r-spatial.github.io/sf/reference/geos_binary_pred.html). Next, [*length()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/length) of Base R is used to calculate numbers of pre-schools that fall inside each planning subzone.

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

**Warning**: You should not confuse with [st_intersection()](https://r-spatial.github.io/sf/reference/geos_binary_ops.html).

You can check the summary statistics of the newly derived *PreSch Count* field by using *summary()* as shown in the code chunk below.

```{r}
summary(mpsz3414$`PreSch Count`)
```

To list the planning subzone with the most number of pre-school, the [*top_n()*](https://dplyr.tidyverse.org/reference/top_n.html) of **dplyr** package is used as shown in the code chunk below.

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

**DIY: Calculate the density of pre-school by planning subzone.**

The solution:

Firstly, the code chunk below uses [*st_area()*](https://r-spatial.github.io/sf/reference/geos_measures.html) of **sf** package to derive the area of each planning subzone.

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

Next, [*mutate()*](https://dplyr.tidyverse.org/reference/mutate.html) of [**dplyr**](https://dplyr.tidyverse.org/) package is used to compute the density by using the code chunk below.

```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## Explorotary Data Analysis (EDA)

In practice, many geospatial analytics start with Exploratory Data Analysis. In this section, you will learn how to use appropriate [ggplot2](https://ggplot2.tidyverse.org/) functions to create functional and yet truthful statistical graphs for EDA purposes.

Firstly, we will plot a histogram to reveal the distribution of `PreSch Density`. Conventionally, *hist()* of R Graphics will be used as shown in the code chunk below.

```{r}
hist(mpsz3414$`PreSch Density`)
```

Although the syntax is very easy to use however the output is far from meeting publication quality. Furthermore, the function has limited room for further customisation.

In the code chunk below, appropriate **ggplot2** functions will be used.

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

**DIY: Using ggplot2 method, plot a scatterplot showing the relationship between Pre-school Density and Pre-school Count.**

```{r}


ggplot(data=mpsz3414,

aes(x= as.numeric(`PreSch Density`), y= as.numeric(`PreSch Count`)))+

geom_point() +

labs(title = "Are pre-school even distributed in Singapore?",

subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",

x = "Pre-school density (per km sq)",

y = "Pre-school Count")
```

## 

## 2  Choropleth Mapping with R

## Overview

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

In this chapter, you will learn how to plot functional and truthful choropleth maps by using an R package called \*\*tmap\*\* package.

## Getting Started

In this hands-on exercise, the key R package use is [**tmap**](https://cran.r-project.org/web/packages/tmap/) package in R. Beside **tmap** package, four other R packages will be used. They are:

-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,

-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,

-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and

-   [**sf**](https://cran.r-project.org/web/packages/sf/) for handling geospatial data.

Among the four packages, **readr**, **tidyr** and **dplyr** are part of **tidyverse** package.

The code chunk below will be used to install and load these packages in RStudio.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

### Importing Data into R

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

### Importing Attribute Data into R

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### Data Preparation

#### Data wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### Joining the attribute data and geospatial data

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

### Choropleth Mapping Geospatial Data Using *tmap*

#### Plotting a choropleth map quickly by using *qtm()*

```{r}
tmap_mode("plot")
```

```{r}
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

#### Creating a choropleth map by using *tmap*'s elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### Drawing a choropleth map using *tm_polygons()*

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

#### Drawing a choropleth map using *tm_fill()* and \*tm_border()\*\*

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

### Data classification methods of **tmap**

#### 

#### Plotting choropleth maps with built-in classification methods

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#2 class

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#6 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#10 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#20 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### Plotting choropleth map with custome break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### Colour Scheme

#### Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

### Map Layouts

#### Map Legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

#### Map style

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### Cartographic Furniture

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

```{r}
tmap_style("white")
```

### Drawing Small Multiple Choropleth Maps

#### By assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in *tm_facets()*

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with *tmap_arrange()*

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mappping Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
